A primer to numerical simulations: The perihelion motion of Mercury

### 2018-02-01 tex/MercuryNumerical.tex
Added flowchart, adjusted code, changed footnote, removed \todo package

\usepackage[prependcaption, textsize=tiny]{todonotes}

@@

The observed perihelion motion of Mercury is nowadays determined as
\[(574.10\pm 0.65)'' \ \mbox{per 100 earth years \cite{RevModPhys.19.361}}\]
where the symbol $''$ denotes "arc seconds": 1$''={(1/3600)}^o$. The bulk of this number
can be understood by the presence of other planets within the Newtonian theory, since
their gravitational force also acts on Mercury. However, a
residual motion of
\begin{equation}
\delta \Theta_M = (42.56\pm 0.94)'' \ \mbox{per 100 earth years \cite{RevModPhys.19.361}} \label{delT}
\end{equation}
remained unexplained, until Einstein quantified the predictions of General Relativity to this particular observable \cite{Einstein}
\begin{equation}
\delta \Theta_{GR} = (43.03\pm 0.03)'' \ \mbox{per 100 earth years \cite{RevModPhys.19.361}} \label{delT} \, .
\end{equation}

@@

So far we only fixed the length of the vectors. Next, we set up the initial directions, which will describe the motion in space.
We will build on the existing \python{} module \vpython{}, which provides an implementation for treating vectors as well as their visualization.
The first object of interest is a \texttt{vector}, which takes three-dimensional coordinates as its input.
With our choice of initial conditions (we picked the initial vectors in the perihelion), the velocity of Mercury is perpendicular to the vector which connects Mercury and Sun (see figure~\ref{fig:sun_merc}):
\begin{lstlisting}
# Import the class vector from vpython
from vpython import vector
# Initialize distance and velocity vectors
vec_rM0 = vector(0, rM0, 0)
vec_vM0 = vector(vM0, 0, 0)
\end{lstlisting}
We use the estimate of Eq.~(\ref{eq:check}) as a guidance to fix the time step $\Delta t$ as
\begin{lstlisting}
# Definition of the time step
dt = 2 * vM0 / c_a / 20
\end{lstlisting}
Here the factor $1/20$ makes sure that $\Delta t$ is indeed consistent with Eq.~(\ref{eq:check}).
Now we are in the position to calculate location and velocity of the planet at $t_0+\Delta t$ using the following expressions\footnote{%
	Note, that there are different notions for numerically integrating differential equations with different accuracies.
	Thus, the ordering and exact expressions for updating $a_{MS}$, $v_M$ and $r_{MS}$ are not unique. However, all correct prescriptions lead to the same result for sufficiently small $\Delta t$.
	For this particular problem, the definitions we show are in good balance between being numerically stable and inexpensive to compute.
}%
\begin{lstlisting}
# Compute the strength of the acceleration
temp = 1 + alpha * rS / vec_rM_old.mag + beta * rL2 / vec_rM_old.mag**2
aMS  = c_a * temp / vec_rM_old.mag**2
# Multiply by the direction
vec_aMS = - aMS * ( vec_rM_old / vec_rM_old.mag )
# Update velocity vector
vec_vM_new = vec_vM_old + vec_aMS * dt
# Update position vector
vec_rM_new = vec_rM_old + vec_vM_new * dt
\end{lstlisting}
Note basic vector operations are already implemented in the predefined \texttt{vector} class.
The difference and sum of two vectors, or the scalar vector multiplication return vectors themselves.
Also the magnitude of a vector --- \code{vector.mag} --- is an attribute of the vector and can be easily extracted.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\clearpage
\section{The code covered by the basic course (visualization of the trajectories)}\label{sec:code}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
Below we provide the complete code for part 1 as developed above.
\lstinputlisting[language=Python,basicstyle=\linespread{1}\scriptsize\ttfamily]{mercury_orbit_base.py}
Further examples and template files can be found online \cite{scripts}.

>>>>

\usepackage{tikz}
\usetikzlibrary{shapes.geometric, arrows}
%
\tikzstyle{startstop} = [
	rectangle,
	rounded corners,
	minimum width=3cm,
	minimum height=1cm,
	text width=4.5cm,
	text centered,
	draw=black,
	fill=green!30
]

\tikzstyle{process} = [
	rectangle,
	minimum width=3cm,
	minimum height=1cm,
	text width=4.5cm,
	text centered,
	draw=black,
	fill=blue!25
]
\tikzstyle{decision} = [
	diamond,
	minimum width=4.5cm,
	minimum height=1cm,
	text width=1.7cm,
	text centered,
	draw=black,
	fill=red!30
]
\tikzstyle{arrow} = [thick,->,>=stealth]

@@

The observed perihelion motion of Mercury is nowadays determined as
\[(574.10\pm 0.65)'' \ \mbox{per 100 earth years \cite{RevModPhys.19.361}}\]
where the symbol $''$ denotes "arc seconds": 1$''={(1/3600)}^o$. The bulk of this number
can be understood by the presence of other planets within the Newtonian theory, since
their gravitational force also acts on Mercury. However, a
residual motion of
\begin{equation}
\delta \Theta_M = (42.56\pm 0.94)'' \ \mbox{per 100 earth years \cite{RevModPhys.19.361}} \label{delT}
\end{equation}
remained unexplained, until Einstein quantified the predictions of General Relativity to this particular observable \cite{Einstein}
\begin{equation}
\delta \Theta_{GR} = (43.03\pm 0.03)'' \ \mbox{per 100 earth years \cite{RevModPhys.19.361}} \label{delTGR} \, .
\end{equation}

@@

So far we only fixed the length of the vectors. Next, we set up the initial directions, which will describe the motion in space.
We will build on the existing \python{} module \vpython{}, which provides an implementation for treating vectors as well as their visualization.
The first object of interest is a \texttt{vector}, which takes three-dimensional coordinates as its input.
With our choice of initial conditions (we picked the initial vectors in the perihelion), the velocity of Mercury is perpendicular to the vector which connects Mercury and Sun (see figure~\ref{fig:sun_merc}):
\begin{lstlisting}
# Import the class vector from vpython
from vpython import vector
# Initialize distance and velocity vectors
vec_rM0 = vector(0, rM0, 0)
vec_vM0 = vector(vM0, 0, 0)
\end{lstlisting}\footnote{Note that the code is designed for \vpython{} versions \texttt{7} and later. For \vpython{} versions prior \texttt{7}, e.g., the statement from \code{vpython import vector}  must be replaced by \code{from visual import vector}.}
We use the estimate of Eq.~(\ref{eq:check}) as a guidance to fix the time step $\Delta t$ as
\begin{lstlisting}
# Definition of the time step
dt = 2 * vM0 / c_a / 20
\end{lstlisting}
Here the factor $1/20$ makes sure that $\Delta t$ is indeed consistent with Eq.~(\ref{eq:check}).
Now we are in the position to calculate location and velocity of the planet at $t_0+\Delta t$ using the following expressions\footnote{%
	Note, that there are different notions for numerically integrating differential equations with different accuracies.
	Thus, the ordering and exact expressions for updating $a_{MS}$, $v_M$ and $r_{MS}$ are not unique. However, all correct prescriptions lead to the same result for sufficiently small $\Delta t$.
	For this particular problem, the definitions we show are in good balance between being numerically stable and inexpensive to compute.
}%
\begin{lstlisting}
# Compute the strength of the acceleration
temp = 1 + alpha * rS / vec_rM_old.mag + beta * rL2 / vec_rM_old.mag**2
aMS  = c_a * temp / vec_rM_old.mag**2
# Multiply by the direction
vec_aMS = - aMS * ( vec_rM_old / vec_rM_old.mag )
# Update velocity vector
vec_vM_new = vec_vM_old + vec_aMS * dt
# Update position vector
vec_rM_new = vec_rM_old + vec_vM_new * dt
\end{lstlisting}
Note basic vector operations are already implemented in the predefined \texttt{vector} class.
The difference and sum of two vectors, or the scalar vector multiplication return vectors themselves.
Also the magnitude of a vector --- \code{vector.mag} --- is an attribute of the vector and can be easily extracted.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\clearpage
\section{The code covered by the basic course (visualization of the trajectories)}\label{sec:code}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
Below we provide the complete code as well as a flowchart for part 1 as developed above.
\begin{figure}[htb]
\centering
\begin{tikzpicture}[node distance=1.8cm, every node/.style={scale=0.8}]
  \node (start)  [startstop] {Start};
  \node (proc1)  [process, below of=start] {Definition of\\ physical parameters};
  \draw [arrow]  (start) -- (proc1);
  \node (proc2)  [process, below of=proc1] {Initialize distance and velocity vectors};
  \draw [arrow]  (proc1) -- (proc2);
  \node (proc4)  [process, below of=proc2] {Activate drawing of trajectory};
  \draw [arrow]  (proc2) -- (proc4);
  \node (proc5)  [process, below of=proc4] {Definition of time step \texttt{dt}};
  \draw [arrow]  (proc4) -- (proc5);
  \node (proc6)  [process, below of=proc5] {Set \texttt{t=0}};
  \draw [arrow]  (proc5) -- (proc6);
  \node (proc6b) [process, below of=proc6] {Set \texttt{alpha} and \texttt{beta}};
  \draw [arrow]  (proc6) -- (proc6b);
  \node (dec1)   [decision, below of=proc6b, yshift=-0.7cm] {\texttt{t < Tmax}};
  \draw [arrow]  (proc6b) -- (dec1);
  \node (proc7)  [process, below of=dec1, yshift=-1.2cm] {Update Mercury trajectory};
  \draw [arrow]  (dec1) -- node[anchor=west] {YES} (proc7) ;
  \node (proc8)  [process, below of=proc7] {Calculate \texttt{vec\_aMS(t)}};
  \draw [arrow]  (proc7) -- (proc8);
  \node (proc9)  [process, below of=proc8] {Calculate \texttt{vec\_vM(t+dt)}};
  \draw [arrow]  (proc8) -- (proc9);
  \node (proc10) [process, below of=proc9] {Calculate \texttt{vec\_rM(t+dt)}};
  \draw [arrow]  (proc9) -- (proc10);
  \node (stop)   [startstop, right of=proc7, xshift=3.5cm] {Stop};
  \node (proc11) [process, left of=proc7, xshift=-3.5cm] {\texttt{t=t+dt}};
  \draw [arrow]  (dec1) -| node[anchor=north, xshift=-1.2cm] {NO} (stop);
  \draw [arrow]  (proc10) -| (proc11);
  \draw [arrow]  (proc11) |- (dec1);
\end{tikzpicture}
\caption{\label{fig:flow}Flowchart demonstrating the logical ordering of the example code.}
\end{figure}

\begin{figure}[htb]
\lstinputlisting[language=Python,basicstyle=\linespread{1}\footnotesize\ttfamily]{mercury_orbit_base.py}
\end{figure}
Further examples and template files can be found online \cite{scripts}.
